Density-dependent selection and the population ecology of relative fitness

### 2017-11-28 manuscript.tex
l

The issue can be expressed more formally by revisiting MacArthur's famous analysis of selection in crowded populations \citep{macarthur_1967}. MacArthur considers a population consisting of two types with densities $n_1$ and $n_2$ subject to density-dependent population growth described by
\begin{equation}
\frac{d n_1}{d t}=f_1(n_1,n_2)\qquad\frac{d n_2}{d t}=f_2(n_1,n_2). \label{eq:macgeneral}
\end{equation}
Other than the type densities, the environment is assumed to remain constant. The functions $f_1$ and $f_2$ must decline to zero if $n_1$ or $n_2$ are sufficiently large, because no population has unlimited resources. This defines the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ in $(n_1,n_2)$ space. The outcome of selection is then determined by the relationship between these nullclines. Specifically, a type will be excluded if its nullcline is completely contained in the region bounded by the other type's nullcline. In other words, for a type to have the possibility of persisting, it must be able to keep growing to higher densities than the other type can tolerate in some region of $(n_1,n_2)$ space (Fig.~\ref{fig:Ksel}a).

At face value, the preceding conclusion does not seem compatible with the constant $N$ assumption of relative fitness models. It is not even clear how to connect relative fitness models and MacArthur's argument, since the former only allows for selection along lines defined by $n_1+n_2=N$ for each $N$ (Fig.~\ref{fig:Ksel}b); the $f_1$ and $f_2$ nullclines are thus not even defined. The incompatibility is particularly apparent in some of the most widely-used models of population growth. MacArthur's argument is often presented using the logistic model and its bi-allelic diploid variants [citations]. Positive selection in crowded populations then requires increases in equilibrium population density --- commonly referred to as ``$K$-selection'' (Fig.~\ref{fig:Ksel}a). The Lotka-Volterra competition model also couples selection in crowded populations to changes in total density $N$ (we return to this in BLAH). Moreover, exploitation competition for consumable resources is potentially a decisive factor in many crowded populations. On a single, homogeneous, limiting resource, the type able to deplete the resource to the lowest equilibrium density excludes the other ($R^*$-rule); except in very special cases, this will also entail differences in saturation density. Thus, it would seem that the ubiquitous constant-$N$, relative fitness description of selection is incompatible with a huge class of population ecological processes occurring in nature and experiments.

\begin{figure}
\centering
\includegraphics[scale=0.8]{Kplot.pdf}
\caption{\label{fig:Ksel} (a) MacArthur's dynamical argument for why ``fitness is $K$'' in crowded environments, illustrated using the logistic model $f_1=r_1(1-\frac{n_1+n_2}{K_1})n_1$ and $f_2=r_2(1-\frac{n_1+n_2}{K_2})n_1$ in Eq. \eqref{eq:macgeneral}. Here $r_1=r_2$ and $K_1>K_2$. (b) The constant-$N$, relative fitness description of selection.}
\end{figure}

@@

In our extension of the classic lottery model (Eq.~\ref{eq:delttot}), we do not restrict ourselves to high propagule densities. Eq. \eqref{eq:lottery} is nonsensical if even a single type has low propagule density ($l_i\ll 1$): type $i$ can win at most $m_i$ territories, yet Eq.~\eqref{eq:lottery} demands $c_i l_i/\sum_j c_j l_j$ of the $U$ unoccupied territories, for any value of $U$. Thus, deviations from the mean propagule composition $l_1,l_2,\ldots,l_G$ are important at low density. The reason for this lies in the fundamental discreteness of biological individuals: types with few propagules depend on the outcome of contests in territories where they have at least one propagule present, not some small fraction of a propagule as would be implied by small $l_i$ in Eq.~\eqref{eq:lottery}.

@@

Comparing Eq. \eqref{eq:master} to Eq. \eqref{eq:lottery}, the classic lottery per-propagule success rate $c_i/\overline{c}L$ has been replaced by three separate terms. The first, $e^{-L}$, accounts for propagules which land alone on unoccupied territories; these territories are won without contest. The second, $R_i c_i/\overline{c}$ represents competitive victories when the $i$ type is a rare invader in a high density population, determining its invasion fitness \citep{metz_1992}. The third term, $A_ic_i/\overline{c}$, represents competitive victories when the $i$ type is abundant. The relative importance of these three terms varies with both the overall propagule density $L$ and the relative propagule frequencies $m_i/M$. If $l_i\gg 1$ for all types, we recover the classic lottery model (only the $A_ic_i/\overline{c}$ term remains, and $A_i\rightarrow 1/L$). Note that not all unoccupied territories are claimed each iteration, since under Poisson dispersal a fraction $e^{-L}$ remain unoccupied; the total number of territories gained is thus $\Delta_+ N=U(1-e^{-L})$.

@@

\begin{figure}
\centering
\includegraphics[scale=0.8]{simulationcomparison.pdf}
\caption{\label{fig:simcomp} Comparison of mean field approximation Eq. \eqref{eq:master} with simulations. Per-propagule success probability $\Delta_+ n_i/l_i U$ from Eq. \eqref{eq:master} and its three components, the classic lottery model, and  individual-based simulations of random dispersal and lottery competition. Two types are present, a rare type with $c_1=1.5$, and a common type with $c_2=1$. Simulation points are almost invisible in for the common type due to near exact agreement with Eq. \eqref{eq:master}. Dashed lines in show the breakdown of the classic lottery model. Parameters: $m_1=10^4$ and $m_2=9\times 10^4$ and $U$ is varied between $5\times 10^3$ and $10^6$.} 
\end{figure}

@@

As shown in the Introduction, MacArthur's argument revolves around the behaviour of  ecological models of the general form Eq.~\ref{eq:macgeneral}, which depends on the relationship between the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$. To formalize this relationship, MacArthur used the symbol ``$K$'' to label the four intersection points of the nullclines with the $n_1$ and $n_2$ axes, specifically $f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(0,K_{22})=0$ and $f_2(K_{21},0)=0$. These $K$ values determine whether a region of higher-density growth exists for each type, provided that the nullclines are close to being straight lines. Note that, while $K_{11}$ and $K_{22}$ are saturation densities akin to the $K$ parameter in the logistic model, this is not true for $K_{12}$ and $K_{21}$, which are related to competition between types. For instance, in the Lotka-Volterra competition model we have $f_1(n_1,n_2)=r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1$ where $\alpha_{11}=1/K_{11}$ measures competitive effects within the first type, whereas $\alpha_{12}=1/K_{12}$ measures competitive effects on the first type due to the second. Thus, when MacArthur concludes that  ``fitness is $K$'' in crowded populations \citep[pp. 149]{macarthur_1967}, the meaning is that selection in crowded populations either favors the ability to keep growing at ever higher densities (moving a type's own nullcline outwards), or the ability to suppress the growth of competitors at lower densities (moving the nullcline of competitors inwards). This general idea applies even if the nullclines are nonlinear to such an extent that the ``$K$'' values themselves do not give much information about the regions of high-density growth.

\begin{figure}
\centering
\includegraphics[scale=0.8]{LVvslottery.pdf}
\caption{\label{fig:Ksel} Blah}
\end{figure}

The distinction between

The nullclines of Eq.~\eqref{eq:master}

>>>>

The issue can be expressed more formally by revisiting MacArthur's analysis of selection in crowded populations \citep{macarthur_1967}. MacArthur considers a population consisting of two types with densities $n_1$ and $n_2$ subject to density-dependent population growth described by
\begin{equation}
\frac{d n_1}{d t}=f_1(n_1,n_2)\qquad\frac{d n_2}{d t}=f_2(n_1,n_2). \label{eq:macgeneral}
\end{equation}
Other than densities, the environment is assumed to remain constant. The functions $f_1$ and $f_2$ must decline to zero if $n_1$ or $n_2$ are sufficiently large, because no population has unlimited resources. This defines the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ in $(n_1,n_2)$ space. The outcome of selection is then determined by the relationship between these nullclines. Specifically, a type will be excluded if its nullcline is completely contained in the region bounded by the other type's nullcline. In other words, for a type to have the possibility of persisting, it must be able to keep growing to higher densities than the other type can tolerate in some region of $(n_1,n_2)$ space (Fig.~\ref{fig:Ksel}a).

This conclusion about selection in crowded populations does not seem compatible with the constant-$N$ assumption of relative fitness models. The incompatibility is particularly apparent in some of the most widely-used models of density-regulated population growth (to illustrate our point we stick to the simple haploid case): 1) The logistic model: the type with the greatest saturation population density excludes the others. 2) The ``$R^*$ rule'', a central tenet of resource competition theory, states that when multiple types are exploiting a homogeneous consumable resource that limits their growth, the type able to deplete the resource to the lowest equilibrium density $R^*$ excludes the others \citep{grover_1997}. Except in special cases, differences in $R^*$ entails differences in saturation population density. 3) The Lotka-Volterra competition model also couples selection in crowded populations to changes in total density $N$ (we return to this in BLAH). Thus, it would seem that the ubiquitous constant-$N$, relative fitness description of selection is incompatible with a huge class of population ecological processes occurring in nature and experiments.

It is not even clear how to connect relative fitness models to density-regulated growth, since they only describe selection along lines defined by $n_1+n_2=N$ for each $N$ (Fig.~\ref{fig:Ksel}b); the $f_1$ and $f_2$ nullclines are thus not defined.

\begin{figure}
\centering
\includegraphics[scale=0.8]{Kplot.pdf}
\caption{\label{fig:Ksel} (a) MacArthur's dynamical argument for how selection operates in crowded environments, illustrated using the logistic model $f_1=r_1(1-\frac{n_1+n_2}{K_1})n_1$ and $f_2=r_2(1-\frac{n_1+n_2}{K_2})n_1$ in Eq. \eqref{eq:macgeneral}. Here $r_1=r_2$ and $K_1>K_2$. (b) The constant-$N$, relative fitness description of selection.}
\end{figure}

@@

Eq. \eqref{eq:lottery} breaks down for types with low propagule density ($l_i\ll 1$) because territorial acquisition is then not correctly represented by a lottery in each territory with the mean propagule density. Instead, a rare type's propagules only make it to a few territories where at least one of its propagule present. In our extension of the classic lottery model, we correct (Eq.~\ref{eq:delttot}) to account for this.

%type $i$ can win at most $m_i$ territories, yet Eq.~\eqref{eq:lottery} demands $c_i l_i/\sum_j c_j l_j$ of the $U$ unoccupied territories, for any value of $U$.

@@

Comparing Eq. \eqref{eq:master} to Eq. \eqref{eq:lottery}, the classic lottery per-propagule success rate $c_i/\overline{c}L$ has been replaced by three separate terms. The first, $e^{-L}$, accounts for propagules which land alone on unoccupied territories; these territories are won without contest. The second, $R_i c_i/\overline{c}$ represents competitive victories when the $i$ type is a rare invader in a high density population, determining its invasion fitness \citep{metz_1992}. The third term, $A_i c_i/\overline{c}$, represents competitive victories when the $i$ type is abundant. The relative importance of these three terms varies with both the overall propagule density $L$ and the relative propagule frequencies $m_i/M$. If $l_i\gg 1$ for all types, we recover the classic lottery model (only the $A_ic_i/\overline{c}$ term remains, and $A_i\rightarrow 1/L$). Note that not all unoccupied territories are claimed each iteration, since under Poisson dispersal a fraction $e^{-L}$ remain unoccupied; the total number of territories gained is thus $\Delta_+ N=U(1-e^{-L})$.

@@

\begin{figure}
\centering
\includegraphics[scale=0.8]{simulationcomparison.pdf}
\caption{\label{fig:simcomp} Comparison of mean field approximation Eq. \eqref{eq:master} with simulations. Per-propagule success probability $\Delta_+ n_i/l_i U$ from the classic lottery model, individual-based simulations of random dispersal and lottery competition, and Eq. \eqref{eq:master} and its three components. Two types are present, a rare type with $c_1=1.5$, and a common type with $c_2=1$. Simulation points are almost invisible in for the common type due to near exact agreement with Eq. \eqref{eq:master}. Dashed lines in show the breakdown of the classic lottery model. Parameters: $m_1=10^4$ and $m_2=9\times 10^4$ and $U$ is varied between $5\times 10^3$ and $10^6$.} 
\end{figure}

@@

--- commonly referred to as ``$K$-selection'' (Fig.~\ref{fig:Ksel}a)

As shown in the Introduction, MacArthur's argument revolves around the behaviour of  ecological models of the general form Eq.~\ref{eq:macgeneral}, which depends on the relationship between the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$. To formalize this relationship, MacArthur used the symbol ``$K$'' to label the four intersection points of the nullclines with the $n_1$ and $n_2$ axes, specifically $f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(0,K_{22})=0$ and $f_2(K_{21},0)=0$. These $K$ values determine whether a region of higher-density growth exists for each type, provided that the nullclines are close to being straight lines. Note that only $K_{11}$ and $K_{22}$ are saturation densities akin to the $K$ parameter in the logistic model; $K_{12}$ and $K_{21}$ are related to competition between types. For instance, in the Lotka-Volterra competition model we have $f_1(n_1,n_2)=r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1$ where $\alpha_{11}=1/K_{11}$ measures competitive effects within the first type, whereas $\alpha_{12}=1/K_{12}$ measures competitive effects on the first type due to the second (Fig.~\ref{fig:LVvslottery}a). Thus, when MacArthur concludes that  ``fitness is $K$'' in crowded populations \citep[pp. 149]{macarthur_1967}, the meaning is that selection either favors the ability to keep growing at ever higher densities (moving a type's own nullcline outwards), or the ability to suppress the growth of competitors at lower densities (moving the nullcline of competitors inwards). This general idea applies even if the nullclines are nonlinear to such an extent that the ``$K$'' values themselves do not give much information about the regions of high-density growth.

\begin{figure}
\centering
\includegraphics[scale=0.8]{LVvslottery.pdf}
\caption{\label{fig:LVvslottery} Blah}
\end{figure}

In Fig.~\ref{fig:LVvslottery} we show a phase diagram for Eq.~\eqref{eq:master}
